---
title: "Creating Tables with tfl_table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Tables with tfl_table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

`writetfl` can render data frames as paginated, publication-quality tables inside
multi-page PDFs. The two key functions are `tfl_table()`, which builds a table
configuration object from a data frame, and `export_tfl()`, which renders it to a
PDF file.

```{r load}
library(writetfl)
library(dplyr)   # for group_by() examples
```

---

## Basic usage

Pass any data frame to `tfl_table()`, then hand the result to `export_tfl()`.
Page annotations, page dimensions, and the output file all belong to
`export_tfl()` — not to `tfl_table()`.

```{r basic, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(head(mtcars, 20))

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Table 1. Motor Trend Car Road Tests (first 20 rows)",
  footer_left = "Source: Motor Trend (1974)",
  header_rule = TRUE
)
```

`tfl_table()` returns a table configuration object. It does not draw anything or
open any device. All drawing happens inside `export_tfl()`.

---

## Column labels

By default, column names are used as column headers. Supply `col_labels` to
override them — either as a named character vector (match by column name) or as a
positional vector the same length as `cols`.

```{r col-labels, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Subset columns and rename them for the report
tbl <- tfl_table(
  head(mtcars, 20)[, c("mpg", "cyl", "hp", "wt")],
  col_labels = c(
    mpg = "Miles/Gallon",
    cyl = "Cylinders",
    hp  = "Horsepower",
    wt  = "Weight\n(1000 lb)"   # \n produces a two-line header
  )
)

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Table 1. Selected Performance Metrics"
)
```

Embedding `\n` in a label creates a multi-line column header. The header row
height is sized automatically to fit the tallest label.

---

## Column widths

Three width modes are available and can be mixed freely within the same table.

| Mode | How to specify | Effect |
|------|----------------|--------|
| Fixed | `unit(1.5, "inches")` | Always exactly that width |
| Relative | Plain numeric, e.g. `2` | Width proportional to remaining space after fixed columns are placed |
| Auto | `NULL` (the default) | Sized to the widest content, up to a per-column maximum |

```{r col-widths, eval = FALSE}
tbl <- tfl_table(
  head(mtcars, 20)[, c("mpg", "cyl", "hp", "wt", "gear")],
  col_widths = list(
    mpg  = unit(1.2, "inches"),   # fixed
    cyl  = 1,                     # relative (equal share of remaining space)
    hp   = 1,                     # relative (equal share)
    wt   = unit(1.4, "inches"),   # fixed
    gear = NULL                   # auto: sized to content
  )
)

export_tfl(tbl, file = "column_widths.pdf")
```

When relative widths are used, they are scaled proportionally among themselves
after all fixed and auto columns have claimed their space.

---

## Column alignment

Numeric columns default to right-aligned; character columns default to
left-aligned. Override per column with `col_align`.

```{r col-align, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ae_summary <- data.frame(
  system_organ_class = c("Gastrointestinal", "Nervous system", "Skin"),
  n_subjects         = c(12L, 7L, 4L),
  pct                = c(24.0, 14.0, 8.0),
  stringsAsFactors   = FALSE
)

tbl <- tfl_table(
  ae_summary,
  col_labels = c(
    system_organ_class = "System Organ Class",
    n_subjects         = "n",
    pct                = "(%)"
  ),
  col_align = c(
    system_organ_class = "left",
    n_subjects         = "right",
    pct                = "right"
  )
)

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Table 2. Adverse Events by System Organ Class",
  footnote    = "Percentages are based on the safety population (N = 50)."
)
```

Valid alignment values are `"left"`, `"right"`, and `"center"`.

---

## Row grouping

Mark grouping columns with `dplyr::group_by()` before passing the data to
`tfl_table()`. Group columns must appear first in the column list; they act as
row headers and their values are suppressed on repeated consecutive rows, giving
the indented-group appearance common in clinical tables.

```{r grouping, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Demographic summary with visit and treatment group as row headers
pk_data <- data.frame(
  visit     = rep(c("Week 4", "Week 8", "Week 12"), each = 4),
  treatment = rep(c("Placebo", "Active 10 mg", "Active 20 mg", "Active 40 mg"), 3),
  n         = c(48L, 50L, 49L, 51L, 45L, 47L, 48L, 50L, 41L, 43L, 44L, 46L),
  mean_auc  = c(120.4, 145.2, 178.9, 201.3,
                118.7, 148.6, 185.2, 219.4,
                115.1, 152.3, 191.7, 228.6),
  sd_auc    = c(18.2, 22.4, 27.6, 31.1,
                17.9, 23.1, 28.4, 32.7,
                17.1, 24.0, 29.2, 34.3),
  stringsAsFactors = FALSE
)

tbl <- pk_data |>
  group_by(visit) |>
  tfl_table(
    col_labels = c(
      visit     = "Visit",
      treatment = "Treatment",
      n         = "n",
      mean_auc  = "Mean AUC\n(ng\u00b7h/mL)",
      sd_auc    = "SD"
    ),
    col_widths = list(
      visit     = unit(1.0, "inches"),
      treatment = unit(1.5, "inches"),
      n         = NULL,
      mean_auc  = unit(1.4, "inches"),
      sd_auc    = unit(0.8, "inches")
    )
  )

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Table 3. PK Summary by Visit and Treatment",
  footnote    = "AUC = area under the concentration-time curve."
)
```

The `suppress_repeated_groups` argument (default `TRUE`) controls whether
repeated group values are hidden. Set it to `FALSE` to show every row's group
value explicitly.

Group values within a block are kept together on the same page wherever possible
(see [Multi-page tables] below).

---

## Multi-page tables

### Row pagination

When a table has more rows than fit on one page, `tfl_table` paginates
automatically. Groups are kept together: if the rows belonging to one group value
do not fit on the current page, the whole group moves to the next page.

Continuation markers are appended to the last column header on intermediate pages
and to the first data row of a continuation page so the reader can follow the
table across page breaks.

```{r row-pagination, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- iris |>
  relocate(Species) |>
  group_by(Species) |>
  tfl_table(
    col_labels = c(
      Species      = "Species",
      Sepal.Length = "Sepal\nLength",
      Sepal.Width  = "Sepal\nWidth",
      Petal.Length = "Petal\nLength",
      Petal.Width  = "Petal\nWidth"
    ),
    row_cont_msg = c("(continued from previous page)", "(continued on next page)"),
    col_cont_msg = "(continued)"
  )

export_tfl(
  tbl,
  preview     = c(1, 2),
  header_left = "Table 4. Iris Measurements by Species",
  header_rule = TRUE,
  footer_rule = TRUE
)
```

### Column pagination

If the total column width exceeds the printable area, `tfl_table` splits the
columns across additional pages. Row-header columns (i.e., the grouped columns)
are repeated at the left of every column page so the reader always knows which
group they are in.

```{r col-pagination, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Lab safety panel: 6 parameters × 13 timepoints — too wide for one page.
# Group by parameter so the parameter column repeats as a row-header on each
# column-split page.
lab_wide <- data.frame(
  parameter = c("ALT (U/L)", "AST (U/L)", "ALP (U/L)",
                "Total Bilirubin (mg/dL)", "Creatinine (mg/dL)", "eGFR (mL/min)"),
  scr   = c(28, 22, 74, 0.6, 0.92, 88),
  bl    = c(30, 24, 76, 0.7, 0.93, 87),
  wk2   = c(33, 26, 79, 0.7, 0.95, 85),
  wk4   = c(35, 28, 81, 0.8, 0.97, 83),
  wk6   = c(36, 29, 83, 0.8, 0.99, 81),
  wk8   = c(38, 30, 84, 0.9, 1.01, 79),
  wk12  = c(36, 28, 82, 0.8, 1.02, 78),
  wk16  = c(34, 27, 80, 0.8, 1.04, 77),
  wk20  = c(33, 26, 78, 0.7, 1.05, 76),
  wk24  = c(31, 25, 76, 0.7, 1.07, 75),
  wk28  = c(30, 24, 75, 0.6, 1.08, 74),
  wk32  = c(29, 23, 74, 0.6, 1.09, 73),
  eot   = c(28, 22, 73, 0.6, 1.10, 72),
  stringsAsFactors = FALSE
)

tbl <- lab_wide |>
  group_by(parameter) |>
  tfl_table(
    col_labels = c(
      parameter = "Lab Parameter",
      scr   = "Screen-\ning",
      bl    = "Base-\nline",
      wk2   = "Week 2",
      wk4   = "Week 4",
      wk6   = "Week 6",
      wk8   = "Week 8",
      wk12  = "Week 12",
      wk16  = "Week 16",
      wk20  = "Week 20",
      wk24  = "Week 24",
      wk28  = "Week 28",
      wk32  = "Week 32",
      eot   = "End of\nTreatment"
    )
  )

export_tfl(
  tbl,
  preview     = c(1, 2),
  header_left = "Table 5. Mean Lab Safety Values by Timepoint",
  header_rule = TRUE
)
```

By default `allow_col_split = TRUE`. Set it to `FALSE` if you want an error
rather than an automatic split — useful during development to confirm that your
column widths fit within the target page dimensions.

```{r col-split-error, error = TRUE}
# This will error if the columns are too wide for the page
tbl_no_split <- tfl_table(
  mtcars,
  allow_col_split = FALSE
)

export_tfl(tbl_no_split, preview = TRUE)
```

### Balancing columns across pages

By default the greedy algorithm packs as many columns as possible onto each
page, which can leave the last page looking sparse — for example, 8 columns on
page 1 and only 2 on page 2. Set `balance_col_pages = TRUE` to redistribute
the data columns so that each page carries approximately the same number.

The greedy pass still runs first to determine the minimum number of pages
required. The data columns are then divided as evenly as possible across those
pages (pages that cannot be divided exactly get one extra column on the earlier
pages). Each candidate balanced group is verified to fit within the available
width; if any group would overflow, the greedy layout is used as a fallback.

```{r col-pagination-balanced, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl_balanced <- lab_wide |>
  group_by(parameter) |>
  tfl_table(
    col_labels = c(
      parameter = "Lab Parameter",
      scr   = "Screen-\ning",
      bl    = "Base-\nline",
      wk2   = "Week 2",
      wk4   = "Week 4",
      wk6   = "Week 6",
      wk8   = "Week 8",
      wk12  = "Week 12",
      wk16  = "Week 16",
      wk20  = "Week 20",
      wk24  = "Week 24",
      wk28  = "Week 28",
      wk32  = "Week 32",
      eot   = "End of\nTreatment"
    ),
    balance_col_pages = TRUE
  )

export_tfl(
  tbl_balanced,
  preview     = c(1, 2),
  header_left = "Table 5b. Mean Lab Safety Values by Timepoint (balanced columns)",
  header_rule = TRUE
)
```

---

## Word-wrapping columns

`wrap_cols` accepts a named numeric vector of column indices or names, specifying
the maximum number of characters before wrapping. This is useful for free-text
columns (narrative descriptions, verbatim terms) that would otherwise force
very wide pages or illegible small fonts.

```{r wrap-cols, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ae_verbatim <- data.frame(
  subject_id    = c("001-001", "001-002", "001-003", "002-001", "002-002"),
  ae_term       = c(
    "Nausea and vomiting, mild, considered possibly related to study treatment",
    "Headache, moderate, considered unlikely related",
    "Fatigue, mild, relationship to study drug uncertain",
    "Abdominal pain, moderate, considered probably related",
    "Dizziness, mild, considered possibly related"
  ),
  onset_day     = c(3L, 7L, 2L, 14L, 5L),
  stringsAsFactors = FALSE
)

tbl <- tfl_table(
  ae_verbatim,
  col_labels = c(
    subject_id = "Subject ID",
    ae_term    = "Adverse Event (Verbatim)",
    onset_day  = "Onset\n(Day)"
  ),
  col_widths = list(
    subject_id = unit(0.8, "inches"),
    ae_term    = unit(3.5, "inches"),
    onset_day  = NULL
  ),
  wrap_cols = "ae_term"
)

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Listing 1. Adverse Event Verbatim Terms",
  header_rule = TRUE
)
```

---

## Handling missing values

`na_string` controls how `NA` values are displayed in the table. The default is
`""` (an empty cell). Supply any character string to substitute a visible token.

```{r na-string, fig.width = 11, fig.height = 8.5, out.width = "100%"}
labs_data <- data.frame(
  subject_id = c("001", "001", "002", "002", "003"),
  visit      = c("Baseline", "Week 4", "Baseline", "Week 4", "Baseline"),
  ALT        = c(28, 31, NA, 45, 22),
  AST        = c(19, NA, 24, 38, 17),
  stringsAsFactors = FALSE
)

tbl <- labs_data |>
  group_by(subject_id) |>
  tfl_table(
    col_labels = c(
      subject_id = "Subject",
      visit      = "Visit",
      ALT        = "ALT\n(U/L)",
      AST        = "AST\n(U/L)"
    ),
    na_string = "NC"   # NC = not collected
  )

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Table 6. Laboratory Values",
  footnote    = "NC = not collected."
)
```

---

## Per-column specification with `tfl_colspec()`

For complex tables it can be cleaner to specify each column separately using
`tfl_colspec()` and collect the results into a list. This avoids long parallel
vectors for labels, widths, and alignments.

```{r tfl-colspec, fig.width = 11, fig.height = 8.5, out.width = "100%"}
pk_summary <- data.frame(
  param     = rep(c("Cmax", "AUC0-inf", "t1/2"), each = 3),
  treatment = rep(c("Placebo", "Active 10 mg", "Active 20 mg"), 3),
  geo_mean  = c(0.00, 145.2, 210.8, 0.00, 4820, 7340, 0.00, 8.4, 9.1),
  cv_pct    = c(NA, 28.4, 31.2, NA, 22.7, 25.8, NA, 15.3, 17.9),
  stringsAsFactors = FALSE
)

tbl <- pk_summary |>
  group_by(param) |>
  tfl_table(
    cols = list(
      tfl_colspec("param",     label = "Parameter",    width = unit(1.2, "inches"), align = "left"),
      tfl_colspec("treatment", label = "Treatment",    width = unit(1.5, "inches"), align = "left"),
      tfl_colspec("geo_mean",  label = "Geometric\nMean", width = unit(1.2, "inches"), align = "right"),
      tfl_colspec("cv_pct",    label = "CV%",          width = unit(0.8, "inches"), align = "right")
    ),
    na_string = "--"
  )

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Table 7. PK Parameters — Geometric Mean (CV%)",
  footnote    = c(
    "CV% = coefficient of variation.",
    "-- = not applicable (placebo)."
  )
)
```

`tfl_colspec()` accepts `col`, `label`, `width`, `align`, `wrap`, and `gp`.
It provides no functionality beyond what the parallel-vector approach offers;
the choice is stylistic.

---

## Typography with `gp`

The `gp` argument to `tfl_table()` controls cell typography. Pass a single
`gpar()` for a uniform style, or a named list for per-section control. For a
full reference of all `gp` keys and their effects, see
`vignette("tfl_table_styling")`.

```{r gp, eval = FALSE}
tbl <- tfl_table(
  head(mtcars, 15)[, c("mpg", "cyl", "hp", "wt")],
  col_labels = c(
    mpg = "MPG",
    cyl = "Cylinders",
    hp  = "Horsepower",
    wt  = "Weight"
  ),
  gp = list(
    header = gpar(fontsize = 9, fontface = "bold"),
    body   = gpar(fontsize = 9)
  )
)

export_tfl(
  tbl,
  file = "typed_table.pdf",
  gp   = gpar(fontsize = 9)   # page annotation text
)
```

---

## Page layout and annotations

Page dimensions, margins, header/footer text, separator rules, and page numbering
are all arguments to `export_tfl()`, not `tfl_table()`. This keeps the table
structure independent from the output format, allowing the same `tfl_table`
object to be used with different page layouts.

```{r annotations, fig.width = 8.5, fig.height = 11, out.width = "100%"}
tbl <- tfl_table(
  head(iris, 30),
  col_labels = c(
    Species      = "Species",
    Sepal.Length = "Sepal\nLength",
    Sepal.Width  = "Sepal\nWidth",
    Petal.Length = "Petal\nLength",
    Petal.Width  = "Petal\nWidth"
  )
)

export_tfl(
  tbl,
  preview       = TRUE,
  pg_width      = 8.5,
  pg_height     = 11,
  margins       = unit(c(t = 1, r = 0.75, b = 1, l = 0.75), "inches"),
  header_left   = "Protocol XY-001\nDraft — Not for Distribution",
  header_center = "CONFIDENTIAL",
  header_right  = format(Sys.Date(), "%d %b %Y"),
  caption       = "Table 8. Iris Sepal and Petal Measurements.",
  footnote      = "Data: Fisher (1936). All measurements in centimetres.",
  footer_left   = "Department of Statistics",
  header_rule   = TRUE,
  footer_rule   = TRUE
)
```

See `vignette("figure_output")` for a full reference of all `export_tfl()`
layout arguments, including typography (`gp`), padding, rules, and overlap
detection.

---

## Summary of `tfl_table()` arguments

| Argument | Default | Purpose |
|----------|---------|---------|
| `x` | — | Data frame or grouped tibble |
| `cols` | `NULL` (all columns) | `NULL` or a list of `tfl_colspec()` objects. To display a column subset, pre-select columns in `x` before passing to `tfl_table()`. |
| `col_widths` | `NULL` (auto) | Named list of `unit()`, plain numeric, or `NULL` per column |
| `col_labels` | column names | Named character vector of header labels; supports `\n` |
| `col_align` | type-based | Named vector: `"left"`, `"right"`, or `"center"` |
| `wrap_cols` | `NULL` | Names of columns to word-wrap |
| `min_col_width` | `unit(0.5, "inches")` | Floor applied to auto-sized columns |
| `allow_col_split` | `TRUE` | If `FALSE`, error when columns exceed page width |
| `balance_col_pages` | `FALSE` | If `TRUE`, redistribute columns evenly across column-split pages instead of packing left-to-right |
| `suppress_repeated_groups` | `TRUE` | Hide repeated group values in consecutive rows |
| `col_cont_msg` | `"Columns continue on other pages"` | Footer message on non-final column pages |
| `row_cont_msg` | `c("(continued)", "(continued on next page)")` | `[1]` shown at top of continuation page; `[2]` shown at bottom of page before continuation |
| `show_col_names` | `TRUE` | Whether to render the column header row at all |
| `col_header_rule` | `TRUE` | Rule below column headers |
| `group_rule` | `TRUE` | Rule above each new group block |
| `group_rule_after_last` | `FALSE` | Rule after the last group block |
| `row_header_sep` | `FALSE` | Vertical rule after row-header columns |
| `na_string` | `""` | Replacement for `NA` values |
| `gp` | `list()` | Typography for headers and body cells |
| `cell_padding` | `unit(c(0.2, 0.5), "lines")` | Vertical and horizontal padding inside each cell |
| `line_height` | `1.05` | Inter-line spacing multiplier for word-wrapped cells |
| `max_measure_rows` | `Inf` | Number of rows sampled when measuring auto column widths |
