---
title: "Styling Tables with tfl_table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Styling Tables with tfl_table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r load}
library(writetfl)
library(dplyr)   # for group_by()
```

```{r example-data}
# Clinical data and column spec used throughout this vignette.
# treatment is the first column so it can serve as the group column.
clinical <- data.frame(
  treatment  = c(rep("Active (n=120)", 3),       rep("Placebo (n=118)", 3)),
  subgroup   = c("All patients", "Age < 65",    "Age \u2265 65",
                 "All patients", "Age < 65",    "Age \u2265 65"),
  n          = c(120L,  74L,  46L,  118L,  71L,  47L),
  responders = c( 68L,  44L,  24L,   31L,  18L,  13L),
  rate_pct   = c(56.7, 59.5, 52.2,  26.3, 25.4, 27.7),
  stringsAsFactors = FALSE
)

col_spec <- list(
  tfl_colspec("treatment",  label = "Treatment Arm",  width = unit(1.3, "inches")),
  tfl_colspec("subgroup",   label = "Subgroup",       width = unit(1.4, "inches")),
  tfl_colspec("n",          label = "N",              width = unit(0.4, "inches")),
  tfl_colspec("responders", label = "Resp.",          width = unit(0.5, "inches")),
  tfl_colspec("rate_pct",   label = "Rate (%)",       width = unit(0.65, "inches"))
)
```

---

## 1. Overview

`tfl_table()` accepts a `gp` argument that is a **named list of `gpar()` objects**.
Each key targets a specific visual element of the rendered table. Keys that are
not supplied fall back to sensible clinical defaults; you only need to specify the
elements you want to change.

The full set of recognized keys is:

| Key | Targets | Default |
|-----|---------|---------|
| `gp$table` | Base font for all table text | `gpar(fontsize = 9, fontfamily = "sans")` |
| `gp$header_row` | Column header row text; `fill` sets background color | `gpar(fontface = "bold")` (inherits `gp$table`) |
| `gp$data_row` | Data cell text; `fill` sets background (vector alternates) | inherits `gp$table` |
| `gp$group_col` | Row-header column text | inherits `gp$table` |
| `gp$continued` | Continuation marker text | `gpar(fontface = "italic")` (inherits `gp$table`) |
| `gp$col_header_rule` | Rule drawn under column header row | `gpar(lwd = 1)` |
| `gp$group_rule` | Rules drawn between groups | `gpar(lwd = 0.5, lty = "dotted")` |
| `gp$row_rule` | Rules drawn between data rows | `gpar(lwd = 0.5)` |
| `gp$row_header_sep` | Vertical rule after row-header columns | `gpar(lwd = 0.5)` |

Inheritance is cascading: `gp$data_row` starts from `gp$table`, so setting
`gp$table = gpar(fontsize = 8)` automatically shrinks data cells unless you
explicitly override `gp$data_row`.

Page-level typography — the text in the page header, caption, footnote, and
footer zones — is controlled by the `gp` argument of `export_tfl_page()` /
`export_tfl()`, not by `tfl_table()`'s `gp`.

---

## 2. Base font — `gp$table`

`gp$table` is the typographic root for the whole table. Every other `gp` key
inherits from it unless overridden.

```{r gp-table, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(
  clinical,
  gp = list(
    table = gpar(fontsize = 8, fontfamily = "serif")
  )
)

export_tfl(tbl, preview = TRUE, header_left = "Base font: serif 8pt")
```

Changing `gp$table` propagates to all rows and rules unless you selectively
override a more specific key.

---

## 3. Column header row style — `gp$header_row` and `show_col_names`

The column header row renders the column names (or labels supplied via
`tfl_colspec()`). By default headers are **bold** at the base font size.

```{r gp-header-row, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Custom header: slightly larger, italic instead of bold
tbl <- tfl_table(
  clinical,
  gp = list(
    table      = gpar(fontsize = 9),
    header_row = gpar(fontface = "italic", fontsize = 10)
  )
)

export_tfl(tbl, preview = TRUE, header_left = "Header row: italic 10pt")
```

Set `show_col_names = FALSE` to suppress the header row entirely — useful when
you are stacking multiple `tfl_table` objects on one page and only the first
needs column labels.

```{r show-col-names, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl_no_header <- tfl_table(
  clinical,
  show_col_names = FALSE
)

export_tfl(tbl_no_header, preview = TRUE,
           header_left = "show_col_names = FALSE")
```

---

## 4. Data row style — `gp$data_row`

`gp$data_row` controls the appearance of every non-header, non-group-column
cell. It inherits `gp$table` automatically.

```{r gp-data-row, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(
  clinical,
  gp = list(
    table    = gpar(fontsize = 9),
    data_row = gpar(col = "grey30")
  )
)

export_tfl(tbl, preview = TRUE, header_left = "Data row: grey text")
```

---

## 5. Group column style — `gp$group_col` and per-column override

Row-header (group) columns — those designated via `dplyr::group_by()` — receive
their own style key, `gp$group_col`, which also inherits `gp$table`.

```{r gp-group-col, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Bold group column to distinguish it from data columns
tbl <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols = list(
      tfl_colspec("treatment",  label = "Treatment Arm", width = unit(1.3, "inches")),
      tfl_colspec("subgroup",   label = "Subgroup",      width = unit(1.4, "inches")),
      tfl_colspec("n",          label = "N",             width = unit(0.4, "inches")),
      tfl_colspec("rate_pct",   label = "Rate (%)",      width = unit(0.65, "inches"))
    ),
    gp = list(
      group_col = gpar(fontface = "bold")
    )
  )

export_tfl(tbl, preview = TRUE, header_left = "Group column: bold")
```

To override a **single** group column without touching the others, pass `gp`
directly to `tfl_colspec()`:

```{r colspec-gp, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# The treatment group column gets bold via its tfl_colspec gp;
# any other group columns would stay at the gp$group_col default
tbl <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols = list(
      tfl_colspec("treatment",  label = "Treatment Arm", width = unit(1.3, "inches"),
                  gp = gpar(fontface = "bold")),
      tfl_colspec("subgroup",   label = "Subgroup",      width = unit(1.4, "inches")),
      tfl_colspec("n",          label = "N",             width = unit(0.4, "inches")),
      tfl_colspec("rate_pct",   label = "Rate (%)",      width = unit(0.65, "inches"))
    )
  )

export_tfl(tbl, preview = TRUE,
           header_left = "Per-colspec gp overrides group_col gp")
```

The `gp` on `tfl_colspec()` takes precedence over `gp$group_col` for that
specific column; all other group columns still inherit `gp$group_col`.

---

## 6. Continuation marker style — `gp$continued` and `row_cont_msg`

When a table spans multiple pages, `tfl_table()` injects a continuation marker
at the bottom of each non-final page. By default the marker text is
`"(continued)"` and is rendered in italic.

```{r gp-continued, eval = FALSE}
# Smaller continuation marker, explicit message
tbl <- tfl_table(
  clinical,
  row_cont_msg = "(table continues on next page)",
  gp = list(
    continued = gpar(fontface = "italic", fontsize = 7, col = "grey50")
  )
)
```

`row_cont_msg` replaces the default `"(continued)"` string. `gp$continued`
controls the visual rendering of whatever text `row_cont_msg` provides. The
continuation marker only appears on tables with more rows than fit on one page;
see `vignette("v02-tfl_table_intro")` for an example.

---

## 7. Horizontal rules — `col_header_rule`, `group_rule`, `group_rule_after_last`

Three boolean arguments switch rules on or off; their corresponding `gp` keys
control line appearance.

### Column header rule

A horizontal rule drawn immediately below the column header row.

```{r col-header-rule, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Thicker header rule
tbl <- tfl_table(
  clinical,
  col_header_rule = TRUE,
  gp = list(
    col_header_rule = gpar(lwd = 1.5)
  )
)

export_tfl(tbl, preview = TRUE, header_left = "Header rule: lwd = 1.5")

# No header rule at all
tbl_no_rule <- tfl_table(
  clinical,
  col_header_rule = FALSE
)

export_tfl(tbl_no_rule, preview = TRUE,
           header_left = "col_header_rule = FALSE")
```

### Between-group rules

A rule drawn after each group of rows (defined by changes in the first group
column). `group_rule_after_last` controls whether a rule also appears after the
final group.

```{r group-rule, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Solid thin rules between groups, including after the last one
tbl <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols                  = col_spec,
    group_rule            = TRUE,
    group_rule_after_last = TRUE,
    gp = list(
      group_rule = gpar(lwd = 0.5, lty = "solid")
    )
  )

export_tfl(tbl, preview = TRUE,
           header_left = "Group rules: solid, including after last")

# No between-group rules
tbl_no_grp <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols       = col_spec,
    group_rule = FALSE
  )

export_tfl(tbl_no_grp, preview = TRUE,
           header_left = "group_rule = FALSE")
```

The default `gp$group_rule` is `gpar(lwd = 0.5, lty = "dotted")`. Any valid
`lty` value accepted by `grid` (e.g. `"dashed"`, `"solid"`, `"dotted"`) works
here.

---

## 8. Data row rules — `row_rule` and `gp$row_rule`

A horizontal rule drawn between every pair of consecutive data rows. Enabled
with `row_rule = TRUE`. Unlike `group_rule` (which only appears at group
boundaries), `row_rule` draws a line after every row except the last.

```{r row-rule, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(
  clinical,
  row_rule = TRUE,
  gp = list(
    row_rule = gpar(lwd = 0.3, col = "grey70")
  )
)

export_tfl(tbl, preview = TRUE,
           header_left = "row_rule = TRUE: lines between rows")
```

The default `gp$row_rule` is `gpar(lwd = 0.5)`. Set `row_rule = FALSE`
(the default) to suppress inter-row rules entirely.

---

## 9. Cell background shading — `gp$fill` and `fill_by`

Background colors for the header row and data rows are controlled through
the `fill` field in existing gp keys.

### Header row background

```{r header-fill, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(
  clinical,
  gp = list(
    header_row = gpar(fontface = "bold", fill = "lightblue")
  )
)

export_tfl(tbl, preview = TRUE,
           header_left = "Header row with fill = 'lightblue'")
```

### Alternating row colors (zebra striping)

Pass a vector of colors to `gp$data_row$fill` to alternate between them:

```{r zebra, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(
  clinical,
  gp = list(
    header_row = gpar(fontface = "bold", fill = "steelblue4", col = "white"),
    data_row   = gpar(fill = c("grey95", "white"))
  )
)

export_tfl(tbl, preview = TRUE,
           header_left = "Alternating row colors")
```

### Alternating by group — `fill_by = "group"`

By default, `fill_by = "row"` cycles through the color vector for each data
row. Setting `fill_by = "group"` advances the color index only at group
boundaries, so all rows in the same group share one background color.

```{r fill-by-group, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols    = col_spec,
    fill_by = "group",
    gp = list(
      data_row = gpar(fill = c("grey95", "white"))
    )
  )

export_tfl(tbl, preview = TRUE,
           header_left = "fill_by = 'group': banded groups")
```

---

## 10. Vertical row-header separator — `row_header_sep` and `gp$row_header_sep`

A vertical rule drawn to the right of the last row-header (group) column,
separating the row labels from the data columns. Enabled with
`row_header_sep = TRUE`.

```{r row-header-sep, fig.width = 11, fig.height = 8.5, out.width = "100%"}
# Thin solid vertical separator after the group column
tbl <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols = list(
      tfl_colspec("treatment",  label = "Treatment Arm", width = unit(1.3, "inches")),
      tfl_colspec("subgroup",   label = "Subgroup",      width = unit(1.4, "inches")),
      tfl_colspec("n",          label = "N",             width = unit(0.4, "inches")),
      tfl_colspec("rate_pct",   label = "Rate (%)",      width = unit(0.65, "inches"))
    ),
    row_header_sep = TRUE,
    gp = list(
      row_header_sep = gpar(lwd = 0.75, col = "grey40")
    )
  )

export_tfl(tbl, preview = TRUE,
           header_left = "Row header separator")

# Suppress the separator (default)
tbl_no_sep <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols           = col_spec,
    row_header_sep = FALSE
  )

export_tfl(tbl_no_sep, preview = TRUE,
           header_left = "row_header_sep = FALSE (default)")
```

---

## 11. Cell padding — `cell_padding`

`cell_padding` is a `grid::unit` object that controls the whitespace between
cell content and cell boundaries. It accepts two forms:

**Scalar** — the same padding is applied on all four sides:

```{r cell-padding-scalar, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(
  clinical,
  cell_padding = unit(0.15, "lines")
)

export_tfl(tbl, preview = TRUE, header_left = "Uniform padding: 0.15 lines")
```

**Two-element vector** — separate vertical and horizontal padding. Use this
when you want tighter horizontal spacing but more vertical breathing room:

```{r cell-padding-vh, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- tfl_table(
  clinical,
  cell_padding = unit(c(0.3, 0.1), "lines")   # [1] = vertical, [2] = horizontal
)

export_tfl(tbl, preview = TRUE,
           header_left = "Asymmetric padding: 0.3v / 0.1h lines")
```

The first element controls top and bottom padding; the second controls left and
right. Reducing horizontal padding allows more columns to fit on a page without
reducing font size.

---

## 12. Column continuation message — `col_cont_msg`

When the table has more columns than fit on one page, `tfl_table()` splits
across multiple column-pages. `col_cont_msg` is a character string displayed as
rotated side labels:

- **Clockwise 90°** (reading downward) to the **right** of the table on pages
  where columns continue on a subsequent page.
- **Counter-clockwise 90°** (reading upward) to the **left** of the full table
  (including row-label columns) on pages where columns continue from a prior page.

One line-height of spacing separates the table edge from the text. Set
`col_cont_msg = NULL` to suppress the labels entirely.

```{r col-cont-msg, eval = FALSE}
# Default message
tbl <- tfl_table(
  clinical,
  col_cont_msg = "Columns continue on next page"
)

# Suppress
tbl_no_msg <- tfl_table(
  clinical,
  col_cont_msg = NULL
)
```

---

## 13. Complete example: clinical default vs. publication style

The following pair of examples contrasts the out-of-the-box clinical appearance
with a more compact publication-style variant. Both render using `preview = TRUE`.

### Default clinical look

```{r example-clinical, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl_clinical <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols                  = col_spec,
    col_header_rule       = TRUE,
    group_rule            = TRUE,
    group_rule_after_last = FALSE,
    row_header_sep        = TRUE,
    cell_padding          = unit(0.2, "lines")
    # gp uses built-in defaults: bold headers, dotted group rules, etc.
  )

export_tfl(
  tbl_clinical,
  preview        = TRUE,
  pg_width       = 11,
  pg_height      = 8.5,
  header_left    = "Study XYZ-001",
  header_right   = "CONFIDENTIAL",
  caption        = "Table 1. Response rates by treatment arm and age subgroup.",
  footnote       = "Abbreviations: Resp. = responders; Rate = response rate.",
  footer_left    = "Program: t_resp.R",
  margins        = unit(c(t = 0.75, r = 0.75, b = 0.75, l = 0.75), "inches")
)
```

### Publication style

```{r example-publication, fig.width = 8.5, fig.height = 11, out.width = "100%"}
tbl_publication <- clinical |>
  group_by(treatment) |>
  tfl_table(
    cols                  = col_spec,
    col_header_rule       = TRUE,
    group_rule            = FALSE,   # no between-group rules
    group_rule_after_last = FALSE,
    row_header_sep        = FALSE,   # no vertical separator
    cell_padding          = unit(c(0.25, 0.08), "lines"),
    gp = list(
      # smaller, serif base font
      table           = gpar(fontsize = 8, fontfamily = "serif"),
      # plain (not bold) column headers, slightly larger
      header_row      = gpar(fontface = "plain", fontsize = 9, fontfamily = "serif"),
      # italicized group column
      group_col       = gpar(fontface = "italic"),
      # heavier header rule
      col_header_rule = gpar(lwd = 1.5),
      # smaller, lighter continuation marker
      continued       = gpar(fontface = "italic", fontsize = 7, col = "grey60")
    )
  )

export_tfl(
  tbl_publication,
  preview        = TRUE,
  pg_width       = 8.5,
  pg_height      = 11,
  caption        = "Table 1. Response rates by treatment arm and age subgroup.",
  footnote       = "Resp. = responders; Rate (%) = response rate.",
  margins        = unit(c(t = 1, r = 1, b = 1, l = 1), "inches")
)
```

The two outputs differ visibly in:

- Font family and weight (sans-serif bold headers vs. serif plain headers)
- Presence of group rules and the vertical row-header separator
- Cell padding (uniform vs. asymmetric v/h form)
- Header rule weight
- Continuation marker appearance
