---
title: "Exporting gt Tables to PDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting gt Tables to PDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette covers `export_tfl()` as used with `gt` table objects. For
data-frame tables built with `tfl_table()`, see
`vignette("v02-tfl_table_intro")`. For figure output, see
`vignette("v01-figure_output")`.

```{r load}
library(writetfl)
library(gt)
library(grid)
```

---

## Basic usage

Pass a `gt_tbl` object directly to `export_tfl()`. The title, subtitle,
source notes, and footnotes are automatically extracted and placed in
writetfl's annotation zones (caption and footnote), while the table body is
rendered as a grid grob via `gt::as_gtable()`.

```{r basic, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- gt(head(mtcars, 10)) |>
  tab_header(
    title    = "Motor Trend Car Road Tests",
    subtitle = "First 10 observations"
  ) |>
  tab_source_note("Source: Motor Trend US magazine (1974).")

export_tfl(tbl, preview = TRUE)
```

### Why annotations are extracted

gt normally renders its title, subtitle, source notes, and footnotes inside
the table grob itself. When placed inside writetfl's page layout, this would
cause duplication --- the annotations would appear both in the grob and in
writetfl's header/footer zones. To avoid this, `export_tfl()` extracts gt
annotations into writetfl's annotation fields and strips them from the gt
object before converting to a grob.

The mapping is:

| gt annotation | writetfl field |
|---------------|----------------|
| `tab_header(title = ...)` | `caption` (first line) |
| `tab_header(subtitle = ...)` | `caption` (second line, joined with `\n`) |
| `tab_source_note(...)` | `footnote` |
| `tab_footnote(...)` | `footnote` (combined with source notes) |

---

## Adding page layout elements

All of writetfl's page layout arguments work with gt tables. Pass them via
`...` just as you would for figures.

```{r layout, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- gt(head(iris, 15)) |>
  tab_header(title = "Iris Measurements") |>
  tab_source_note("Source: Fisher (1936).")

export_tfl(
  tbl,
  preview      = TRUE,
  header_left  = "Study Report",
  header_right = format(Sys.Date(), "%d %b %Y"),
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

---

## Multiple gt tables

Pass a list of `gt_tbl` objects to produce a multi-page PDF with one table
per page. Each table's annotations are extracted independently.

```{r multi, eval = FALSE}
tbl1 <- gt(head(mtcars, 10)) |>
  tab_header(title = "Table 1. First 10 rows")

tbl2 <- gt(tail(mtcars, 10)) |>
  tab_header(title = "Table 2. Last 10 rows")

export_tfl(
  list(tbl1, tbl2),
  file         = "two-tables.pdf",
  header_left  = "Appendix",
  header_rule  = TRUE
)
```

---

## Footnotes and source notes

Cell-level footnotes added via `tab_footnote()` and source notes added via
`tab_source_note()` are combined into writetfl's footnote zone.

```{r footnotes, fig.width = 11, fig.height = 8.5, out.width = "100%"}
tbl <- gt(head(mtcars[, 1:6], 8)) |>
  tab_header(title = "Fuel Economy Data") |>
  tab_footnote(
    "Highest in sample.",
    locations = cells_body(columns = mpg, rows = mpg == max(mpg))
  ) |>
  tab_source_note("Source: Motor Trend (1974).")

export_tfl(tbl, preview = TRUE)
```

---

## Automatic pagination for tall tables

When a gt table is too tall to fit on a single page, `export_tfl()` splits
it across multiple pages automatically. Row group boundaries are respected ---
a group is never split across pages.

```{r pagination, eval = FALSE}
# A large table that won't fit on one page
big_data <- data.frame(
  group = rep(c("Treatment A", "Treatment B", "Placebo"), each = 20),
  subject = paste0("SUBJ-", sprintf("%03d", 1:60)),
  value   = round(rnorm(60, 100, 15), 1)
)

tbl <- gt(big_data, groupname_col = "group") |>
  tab_header(title = "Subject-Level Results") |>
  tab_source_note("Source: Clinical Trial XY-001.")

export_tfl(
  tbl,
  file         = "paginated.pdf",
  header_left  = "Study Report",
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

Each page carries the same caption and footnote from the original gt object.
The column headers are repeated on every page since each page is a complete
gt sub-table.

### How pagination works

1. The full table is converted to a grob and its height measured.
2. If it fits within the available content area, a single page is produced.
3. If it overflows, rows are split into groups:
   - **With `tab_row_group()`**: groups of rows defined by `tab_row_group()`
     are kept together. Pages are filled greedily by groups.
   - **Without row groups**: each row is treated as its own unit, allowing
     fine-grained page breaks.
4. For each page chunk, a sub-table is built preserving column labels,
   `fmt_*()` formatting, `tab_style()` styling, column spanners,
   merged columns, and summary rows.

### Ungrouped tables

Tables without explicit row groups are split row-by-row:

```{r ungrouped-pagination, eval = FALSE}
tbl <- gt(mtcars) |>
  tab_header(title = "All 32 Cars") |>
  fmt_number(columns = mpg, decimals = 1)

export_tfl(tbl, file = "all-cars.pdf")
```

---

## Preserved gt features

The following gt features are preserved through pagination:

| Feature | Preserved? | Notes |
|---------|:----------:|-------|
| `tab_header()` | Yes | Extracted as writetfl caption |
| `tab_source_note()` | Yes | Extracted as writetfl footnote |
| `tab_footnote()` | Yes | Combined with source notes |
| `tab_row_group()` | Yes | Groups kept together across pages |
| `tab_spanner()` | Yes | Column spanners repeated on every page |
| `fmt_*()` functions | Yes | Re-indexed per page subset |
| `tab_style()` | Yes | Re-indexed per page subset |
| `cols_merge()` | Yes | Carried through boxhead |
| `summary_rows()` | Yes | Filtered to groups present on each page |
| `cols_label()` | Yes | Carried through boxhead |

```{r spanner-example, eval = FALSE}
tbl <- gt(mtcars[, 1:6]) |>
  tab_header(title = "Motor Trend Data") |>
  tab_spanner(label = "Performance", columns = c(mpg, cyl, disp)) |>
  tab_spanner(label = "Engine", columns = c(hp, drat, wt)) |>
  fmt_number(columns = mpg, decimals = 1) |>
  tab_style(
    style = cell_fill(color = "lightyellow"),
    locations = cells_body(columns = mpg)
  )

export_tfl(tbl, file = "styled.pdf")
```
