---
title: "Exporting rtables Tables to PDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting rtables Tables to PDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette covers `export_tfl()` as used with rtables `VTableTree` objects.
For data-frame tables built with `tfl_table()`, see
`vignette("v02-tfl_table_intro")`. For gt tables, see
`vignette("v05-gt_tables")`. For figure output, see
`vignette("v01-figure_output")`.

```{r load}
library(writetfl)
library(rtables)
library(grid)
```

---

## Basic usage

Pass a `VTableTree` object directly to `export_tfl()`. The main title,
subtitles, main footer, and provenance footer are automatically extracted
and placed in writetfl's annotation zones (caption and footnote), while the
table body is rendered as monospace text via `toString()` and wrapped in a
grid `textGrob`.

```{r basic, fig.width = 11, fig.height = 8.5, out.width = "100%"}
lyt <- basic_table(
  title       = "Iris Sepal Length by Species",
  subtitles   = "Mean values",
  main_footer = "Source: Anderson (1935)."
) |>
  split_cols_by("Species") |>
  analyze("Sepal.Length", mean)

tbl <- build_table(lyt, iris)

export_tfl(tbl, preview = TRUE)
```

### Why annotations are extracted

rtables normally renders its title, subtitles, and footers as part of the
text output. When placed inside writetfl's page layout, this would cause
duplication --- the annotations would appear both in the text rendering and
in writetfl's header/footer zones. To avoid this, `export_tfl()` extracts
rtables annotations into writetfl's annotation fields and strips them from
the rtables object before rendering.

The mapping is:

| rtables annotation | writetfl field |
|--------------------|----------------|
| `main_title` | `caption` (first line) |
| `subtitles` | `caption` (subsequent lines, joined with `\n`) |
| `main_footer` | `footnote` |
| `prov_footer` | `footnote` (appended after main footer) |

---

## Adding page layout elements

All of writetfl's page layout arguments work with rtables tables. Pass them
via `...` just as you would for figures.

```{r layout, fig.width = 11, fig.height = 8.5, out.width = "100%"}
lyt <- basic_table(
  title       = "Iris Measurements",
  main_footer = "Source: Fisher (1936)."
) |>
  split_cols_by("Species") |>
  analyze(c("Sepal.Length", "Sepal.Width"), mean)

tbl <- build_table(lyt, iris)

export_tfl(
  tbl,
  preview      = TRUE,
  header_left  = "Study Report",
  header_right = format(Sys.Date(), "%d %b %Y"),
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

---

## Multiple rtables tables

Pass a list of `VTableTree` objects to produce a multi-page PDF with one
table per page. Each table's annotations are extracted independently.

```{r multi, eval = FALSE}
tbl1 <- basic_table(title = "Table 1: Sepal Length") |>
  split_cols_by("Species") |>
  analyze("Sepal.Length", mean) |>
  build_table(iris)

tbl2 <- basic_table(title = "Table 2: Petal Length") |>
  split_cols_by("Species") |>
  analyze("Petal.Length", mean) |>
  build_table(iris)

export_tfl(
  list(tbl1, tbl2),
  file         = "two-tables.pdf",
  header_left  = "Appendix",
  header_rule  = TRUE
)
```

---

## Automatic pagination

When an rtables table is too tall to fit on a single page, `export_tfl()`
uses rtables' built-in `paginate_table()` to split it across pages. Row
group boundaries are respected --- a group is never split across pages.

```{r pagination, eval = FALSE}
big_data <- data.frame(
  arm     = rep(c("Treatment", "Control"), each = 50),
  site    = rep(paste0("Site ", 1:10), each = 10),
  outcome = rnorm(100, 50, 10)
)

lyt <- basic_table(
  title       = "Subject Outcomes by Site",
  main_footer = "Source: Clinical Trial XY-001."
) |>
  split_cols_by("arm") |>
  split_rows_by("site") |>
  analyze("outcome", mean)

tbl <- build_table(lyt, big_data)

export_tfl(
  tbl,
  file         = "paginated.pdf",
  header_left  = "Study Report",
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

Each page carries the same caption and footnote from the original rtables
object.

### How pagination works

1. The available content height is measured (page height minus margins,
   headers, footers, caption, and footnote).
2. Content dimensions are converted to lines-per-page (`lpp`) and
   characters-per-page (`cpp`) using font metrics.
3. `rtables::paginate_table()` splits the table, respecting row group
   boundaries and rtables' own split rules.
4. Each page is rendered to text via `toString()` and wrapped in a
   `textGrob` with monospace font.

### Font control

The rendering font can be controlled via `...` parameters:

```{r font, eval = FALSE}
export_tfl(
  tbl,
  file                 = "custom-font.pdf",
  rtables_font_family  = "Courier",
  rtables_font_size    = 10,
  rtables_lineheight   = 1.2
)
```

---

## Preserved rtables features

The following rtables features are preserved through the `toString()`
rendering pipeline:

| Feature | Preserved? | Notes |
|---------|:----------:|-------|
| `main_title` | Yes | Extracted as writetfl caption |
| `subtitles` | Yes | Extracted as writetfl caption |
| `main_footer` | Yes | Extracted as writetfl footnote |
| `prov_footer` | Yes | Extracted as writetfl footnote |
| `split_cols_by()` | Yes | Column structure rendered by toString |
| `split_rows_by()` | Yes | Row groups with nesting and indentation |
| `analyze()` | Yes | Analysis rows with formatting |
| `summarize_row_groups()` | Yes | Group summary rows |
| `add_colcounts()` | Yes | Column N counts |
| `append_topleft()` | Yes | Top-left corner label |
| `tab_fn_*()` footnotes | Yes | Referential footnotes |
| Section dividers | Yes | `horizontal_sep`, `section_div` |
| Cell formatting | Yes | All `rcell()` format strings |

### Column splits

```{r col-splits, fig.width = 11, fig.height = 8.5, out.width = "100%"}
lyt <- basic_table(title = "Column Split Example") |>
  split_cols_by("Species") |>
  analyze(c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
          mean)

tbl <- build_table(lyt, iris)
export_tfl(tbl, preview = TRUE)
```

### Row groups with nesting

```{r row-groups, fig.width = 11, fig.height = 8.5, out.width = "100%"}
lyt <- basic_table(title = "Nested Row Groups") |>
  split_rows_by("Species") |>
  analyze(c("Sepal.Length", "Petal.Length"), mean)

tbl <- build_table(lyt, iris)
export_tfl(tbl, preview = TRUE)
```

### Column counts

```{r colcounts, fig.width = 11, fig.height = 8.5, out.width = "100%"}
lyt <- basic_table(title = "With Column Counts") |>
  split_cols_by("Species") |>
  add_colcounts() |>
  analyze("Sepal.Length", mean)

tbl <- build_table(lyt, iris)
export_tfl(tbl, preview = TRUE)
```

### Top-left label

```{r topleft, fig.width = 11, fig.height = 8.5, out.width = "100%"}
lyt <- basic_table(title = "Top-Left Label") |>
  split_cols_by("Species") |>
  append_topleft("Measurement") |>
  analyze(c("Sepal.Length", "Petal.Length"), mean)

tbl <- build_table(lyt, iris)
export_tfl(tbl, preview = TRUE)
```
