---
title: "Exporting flextable Tables to PDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting flextable Tables to PDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette covers `export_tfl()` as used with flextable objects.
For data-frame tables built with `tfl_table()`, see
`vignette("v02-tfl_table_intro")`. For gt tables, see
`vignette("v05-gt_tables")`. For rtables tables, see
`vignette("v06-rtables")`. For figure output, see
`vignette("v01-figure_output")`.

```{r load}
library(writetfl)
library(flextable)
library(grid)
```

---

## Basic usage

Pass a `flextable` object directly to `export_tfl()`. Captions set via
`set_caption()` are automatically extracted and placed in writetfl's caption
zone. Footer rows (from `footnote()` or `add_footer_lines()`) are extracted
into writetfl's footnote zone.

```{r basic, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ft <- flextable(head(iris, 10)) |>
  set_caption("Iris Measurements — First 10 Rows") |>
  add_footer_lines("Source: Anderson (1935).")

export_tfl(ft, preview = TRUE)
```

---

## Caption handling

The caption from `set_caption()` is placed in writetfl's caption section
above the table. It does not appear in the `gen_grob()` output (flextable
reserves captions for document formats like Word and HTML), so there is no
duplication.

```{r caption, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ft <- flextable(head(mtcars[, 1:6], 8)) |>
  set_caption("Table 1. Selected Motor Trend Variables")

export_tfl(ft, preview = TRUE,
  header_left = "Appendix A",
  header_rule = TRUE,
  footer_rule = TRUE
)
```

---

## Footnote extraction

Footer rows added via `footnote()` or `add_footer_lines()` are extracted
as plain text and placed in writetfl's footnote zone below the table. The
footer rows are removed from the flextable before rendering so they don't
appear twice.

When `footnote()` is used, the superscript reference symbols in body cells
are preserved — they still point to the footnotes now positioned in
writetfl's footnote section.

```{r footnotes, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ft <- flextable(head(iris, 8)) |>
  set_caption("Table 2. Iris with Footnotes")

ft <- footnote(ft, i = 1, j = 1, part = "body",
  value = as_paragraph("Measured in centimetres."),
  ref_symbols = "a"
)
ft <- footnote(ft, i = 1, j = 3, part = "body",
  value = as_paragraph("Petal measurements are less variable."),
  ref_symbols = "b"
)

export_tfl(ft, preview = TRUE,
  header_left  = "Study Report",
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

---

## Adding page layout elements

All of writetfl's page layout arguments work with flextable tables. Pass
them via `...` just as you would for figures.

```{r layout, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ft <- flextable(head(mtcars[, 1:6], 10)) |>
  set_caption("Table 3. Motor Trend Cars") |>
  add_footer_lines("Source: Motor Trend (1974).")

export_tfl(
  ft,
  preview      = TRUE,
  header_left  = "Study Report",
  header_right = format(Sys.Date(), "%d %b %Y"),
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

---

## Multiple flextable tables

Pass a list of `flextable` objects to produce a multi-page PDF with one
table per page.

```{r multi, eval = FALSE}
ft1 <- flextable(head(iris, 10)) |>
  set_caption("Table 1. Iris (first 10 rows)")

ft2 <- flextable(tail(iris, 10)) |>
  set_caption("Table 2. Iris (last 10 rows)")

export_tfl(
  list(ft1, ft2),
  file         = "two-tables.pdf",
  header_left  = "Appendix",
  header_rule  = TRUE
)
```

---

## Automatic pagination

When a flextable table is too tall to fit on a single page, `export_tfl()`
splits it across pages by subsetting body rows. The header is repeated on
each page.

```{r pagination, eval = FALSE}
big_ft <- flextable(mtcars[, 1:6]) |>
  set_caption("All Motor Trend Cars") |>
  add_footer_lines("Source: datasets package.")

export_tfl(
  big_ft,
  file         = "paginated.pdf",
  header_left  = "Analysis Report",
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

Each page carries the same caption and footnote from the original
flextable object.

**Note:** When pagination occurs, per-cell formatting applied via
`color()`, `bg()`, `bold()`, etc. is not preserved on paginated pages.
Themes (e.g., `theme_vanilla()`) are not re-applied. For tables with
extensive cell-level formatting, ensure the table fits on a single page
or split the data manually before creating flextable objects.

---

## Preserved features

The following flextable features are preserved through the `gen_grob()`
rendering pipeline:

| Feature | Preserved? | Notes |
|---------|:----------:|-------|
| `set_caption()` | Yes | Extracted as writetfl caption |
| `footnote()` | Yes | Text extracted as writetfl footnote; reference symbols preserved in cells |
| `add_footer_lines()` | Yes | Extracted as writetfl footnote |
| `add_header_row()` | Yes | Rendered as part of table header |
| `add_header_lines()` | Yes | Rendered as part of table header |
| `set_header_labels()` | Yes | Column header labels |
| `merge_v()`, `merge_h()`, `merge_at()` | Yes | Cell merging |
| `border()`, `hline()`, `vline()` | Yes | All border styles |
| `color()`, `bg()` | Yes | Text and background colours |
| `bold()`, `italic()` | Yes | Text emphasis |
| `align()`, `align_text_col()` | Yes | Cell alignment |
| `theme_*()` functions | Yes | All built-in themes |
| `colformat_*()` functions | Yes | Number/date formatting |
| `width()`, `height()` | Yes | Column widths scaled to fit page |
| `as_image()`, `as_raster()` | Partial | Images render but require appropriate device |
| `as_equation()` | No | Equations not supported in grid rendering |
| `hyperlink_text()` | No | Hyperlinks not supported in grid/PDF |

### Themes

```{r themes, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ft <- flextable(head(iris, 8)) |>
  set_caption("Table with Booktabs Theme") |>
  theme_booktabs()

export_tfl(ft, preview = TRUE)
```

### Merged cells

```{r merged, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ft <- flextable(head(iris, 8)) |>
  set_caption("Table with Merged Cells") |>
  merge_v(j = "Species")

export_tfl(ft, preview = TRUE)
```

### Borders and colours

```{r borders, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ft <- flextable(head(mtcars[, 1:5], 6)) |>
  set_caption("Table with Custom Borders") |>
  border_outer(border = officer::fp_border(width = 2)) |>
  bg(i = 1, bg = "#E8F0FE", part = "header")

export_tfl(ft, preview = TRUE)
```
