---
title: "Getting Started with writetfl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with writetfl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r load}
library(writetfl)
library(ggplot2)
library(dplyr)
```

`writetfl` produces multi-page PDF files from `ggplot2` figures, data-frame
tables, and other grid content with precise, composable page layouts required
for clinical trial TFL deliverables and regulatory submissions.

Each page is divided into up to five vertical sections — header, caption,
content, footnote, and footer — whose heights are computed dynamically from
live font metrics so that the content area always fills exactly the remaining
space. Nothing ever overlaps.

---

## Page layout

Every page follows this structure:

```
┌─────────────────────────────────────────────────┐  ← page edge
│              (outer margin)                     │
│  ┌───────────────────────────────────────────┐  │
│  │  header_left  header_center  header_right │  │  header
│  │  ---------------------------------------- │  │  ← header_rule (optional)
│  │  caption                                  │  │  caption
│  │                                           │  │
│  │         content (fills remainder)         │  │  content
│  │                                           │  │
│  │  footnote                                 │  │  footnote
│  │  ---------------------------------------- │  │  ← footer_rule (optional)
│  │  footer_left  footer_center  footer_right │  │  footer
│  └───────────────────────────────────────────┘  │
│              (outer margin)                     │
└─────────────────────────────────────────────────┘
```

Absent sections and their padding gaps are suppressed entirely — no blank space
is reserved for them.

---

## Figures

Pass a single `ggplot` directly. `"Page 1 of 1"` is added to the footer
automatically.

```{r single-figure, fig.width = 11, fig.height = 8.5, out.width = "100%"}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(x = "Weight (1000 lb)", y = "Miles per gallon")

export_tfl(p, preview = TRUE)
```

Build a multi-page report by supplying a list of page specs. Arguments in `...`
are shared across all pages; values inside a page's list element take priority.

```{r multi-page-figure, fig.width = 11, fig.height = 8.5, out.width = "100%"}
pages <- list(
  list(
    content  = ggplot(mtcars, aes(wt, mpg))  + geom_point(),
    caption  = "Figure 1. Weight is negatively associated with fuel efficiency.",
    footnote = "n = 32 vehicles."
  ),
  list(
    content  = ggplot(mtcars, aes(hp, mpg))  + geom_point(),
    caption  = "Figure 2. Higher horsepower predicts lower fuel efficiency.",
    footnote = "Pearson r = -0.78."
  )
)

export_tfl(
  pages,
  preview      = TRUE,
  header_left  = "Fuel Economy Analysis",
  header_right = format(Sys.Date(), "%d %b %Y"),
  header_rule  = TRUE,
  footer_rule  = TRUE
)
```

For the full set of layout controls — separator rules, typography, multi-line
text, overlap detection, preview mode, and more — see
`vignette("v01-figure_output")`.

---

## Data-frame tables

`tfl_table()` converts a data frame into a paginated table grob. Pass the
result directly to `export_tfl()`.

```{r basic-table, fig.width = 11, fig.height = 8.5, out.width = "100%"}
ae_summary <- data.frame(
  system_organ_class = c("Gastrointestinal", "Nervous system", "Skin"),
  n_subjects         = c(12L, 7L, 4L),
  pct                = c(24.0, 14.0, 8.0)
)

tbl <- tfl_table(
  ae_summary,
  col_labels = c(system_organ_class = "System Organ Class",
                 n_subjects = "n", pct = "(%)"),
  col_align  = c(system_organ_class = "left",
                 n_subjects = "right", pct = "right")
)

export_tfl(
  tbl,
  preview     = TRUE,
  header_left = "Table 1. Adverse Events by System Organ Class",
  footnote    = "Percentages are based on the safety population (N = 50)."
)
```

Use `dplyr::group_by()` to designate row-header columns. Group columns repeat
on every column-split page and suppress repeated values in consecutive rows.

```{r grouped-table, fig.width = 11, fig.height = 8.5, out.width = "100%"}
pk_data <- data.frame(
  visit     = rep(c("Week 4", "Week 8", "Week 12"), each = 4),
  treatment = rep(c("Placebo", "Active 10 mg", "Active 20 mg", "Active 40 mg"), 3),
  n         = c(48L, 50L, 49L, 51L, 45L, 47L, 48L, 50L, 41L, 43L, 44L, 46L),
  mean_auc  = c(120.4, 145.2, 178.9, 201.3,
                118.7, 148.6, 185.2, 219.4,
                115.1, 152.3, 191.7, 228.6),
  stringsAsFactors = FALSE
)

pk_data |>
  group_by(visit) |>
  tfl_table(
    col_labels = c(visit = "Visit", treatment = "Treatment",
                   n = "n", mean_auc = "Mean AUC\n(ng\u00b7h/mL)")
  ) |>
  export_tfl(
    preview     = 1,
    header_left = "Table 2. PK Summary by Visit"
  )
```

`tfl_table()` paginates automatically:

- **Row pagination** — rows split across pages with optional `(continued)`
  markers; groups are kept together where possible.
- **Column pagination** — columns that exceed the page width are split across
  pages; set `balance_col_pages = TRUE` to distribute columns evenly.
- **Column widths** — auto-sized from content, fixed (`unit()`), or
  relative-weight numeric.
- **Word wrapping** — `wrap_cols` reflows long text within a fixed column
  width.

For the complete table reference — column specs, continuation messages, cell
padding, line height, and more — see `vignette("v02-tfl_table_intro")`.

For table typography and styling, see `vignette("v03-tfl_table_styling")`.

---

## Multi-page reports

`export_tfl()` accepts a list of page specifications, so different figures can
coexist in one PDF with per-page captions, footnotes, or other annotations
alongside any shared header and footer.

```{r mixed, fig.width = 11, fig.height = 8.5, out.width = "100%"}
export_tfl(
  list(
    list(content = ggplot(mtcars, aes(wt, mpg)) + geom_point(),
         caption  = "Figure 1. Weight vs fuel efficiency.",
         footnote = "Pearson r = -0.87."),
    list(content = ggplot(mtcars, aes(factor(cyl), mpg)) + geom_boxplot() +
                     labs(x = "Cylinders", y = "MPG"),
         caption  = "Figure 2. Fuel efficiency by cylinder count.")
  ),
  preview     = TRUE,
  header_left = "Analysis Report",
  header_rule = TRUE
)
```

`tfl_table()` objects are passed as the top-level `x` argument to
`export_tfl()` rather than inside a page spec list — the function handles
pagination and page construction automatically in that case.

---

## Key shared features

### Automatic page numbering

`page_num` (default `"Page {i} of {n}"`) fills `footer_right` unless a
`footer_right` value is already set. Use a
[glue](https://glue.tidyverse.org/) template or `NULL` to disable.

```{r page-num, eval = FALSE}
export_tfl(pages, file = "numbered.pdf", page_num = "{i} / {n}")
export_tfl(pages, file = "no-numbers.pdf", page_num = NULL)
```

### Typography

Pass a bare `gpar()` to style all annotation text uniformly, or a named list
for section- or element-level control. Resolution priority (highest wins):
element → section → global.

```{r gp, fig.width = 11, fig.height = 8.5, out.width = "100%"}
export_tfl(
  p,
  preview     = TRUE,
  header_left = "Protocol XY-001",
  caption     = "Figure 1. Results.",
  gp = list(
    header  = gpar(fontsize = 11, fontface = "bold"),
    caption = gpar(fontsize =  9, fontface = "italic"),
    footer  = gpar(fontsize =  8)
  )
)
```

### Preview mode

`export_tfl(..., preview = TRUE)` draws to the currently open device without
opening or closing a PDF — useful for interactive layout tuning in RStudio or
Positron, and for inline graphics in vignettes. Pass an integer vector to
render specific pages only.

```{r preview, fig.width = 11, fig.height = 8.5, out.width = "100%"}
export_tfl_page(
  x            = list(content = p),
  header_left  = "Draft",
  caption      = "Figure 1.",
  header_rule  = TRUE,
  preview      = TRUE
)
```

---

## Vignette index

| Vignette | What it covers |
|----------|----------------|
| `vignette("writetfl")` | This overview |
| `vignette("v01-figure_output")` | Full `export_tfl()` / `export_tfl_page()` reference for figures: page dimensions, margins, rules, typography, overlap detection, preview mode |
| `vignette("v02-tfl_table_intro")` | `tfl_table()` in depth: column specs, widths, alignment, wrapping, row/column pagination, group columns |
| `vignette("v03-tfl_table_styling")` | Table typography with `gp`: per-section and per-element `gpar()` overrides, cell padding, line height |
